TADCompare: an R package for differential analysis and characterization of Topologically Associated Domains

### 2019-09-21 manuscript/manuscript.Rmd
Describe Supplementary Figure S1

## Networks of KEGG pathways and randomly selected genes as reference for high and low coherence, respectively

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed negative correlation of network size with the slopes of KEGG and random networks (-0.58/-0.86, respectively, Figure 2, Supplementary Table S2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

Given the dependence of internal and external degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we create size-matched references of high and low coherence. Specifically, we select the 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections) and 100 random networks matched in size for the corresponding network of phenotype-specific genes. Consequently, we use the median slopes of internal vs. external degree distributions created by these size-matched references to normalize the slope of the phenotype-specific network to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed alleviates the dependency between coherence and network size (Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

The potential drawback of this reference size-matching approach is that phenotypes with large network sizes (e.g., "Obesity-related traits", 779 genes) may not have the size-matched reference of high coherence (e.g., the largest KEGG pathways is 374 genes). Consequently, coherence estimation for such networks may be affected by the smaller size of reference of high coherence. Therefore, we use networks from GOCC and REACTOME collections that have larger maximum number of genes (Supplementary Table S2) in parallel with KEGG networks as a reference of high coherence. We found that coherence estimations using KEGG, GOCC, and REACTOME networks as references of high coherence are highly concordant (Average Pearson correlation coefficient > 0.59, Supplementary Figure S2). The use of size-matched networks from KEGG, GOCC, and REACTOME collections enables robust estimation of coherence for phenotype-specific networks of different sizes. Given the similarity of the results obtained with different PPI databases and references of high coherence, we present results obtained with STRING filtered PPI database and KEGG networks as a reference of high coherence, with other observations presented in Supplementary Table S3 and Supplementary Figures S3-S8.

@@

To alleviate the dependency of degree distributions from the size of phenotype-associated gene sets and the content of a selected PPI database, we normalized slopes of phenotype-specific networks to the range of slopes formed by degree distributions of highly coherent and random networks. Specifically, we select the 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections, $\beta_{min}$ median slope) and 100 random networks ($\beta_{max}$ median slope) matched in size for the corresponding network of phenotype-specific genes. For each phenotype $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of KEGG pathways. These values represent normalized coherence and can be compared across phenotypes and PPI databases.

@@

**Supplementary Figure S1. Normalized coherence is unaffected by network size.** Red/blue gradient and numbers represent Pearson correlation coefficients for each pairwise comparison of the number of SNPs, genes, and the normalized coherence estimates using the corresponding PPI databases and KEGG pathways as a reference of high coherence.

>>>>

## Networks of KEGG/REACTOME pathways and GO Cellular Component collection vs. randomly selected genes serve as a reference for high vs. low coherence, respectively

@@

To observe whether network properties affect coherence estimation in experimental settings, we investigated the effect of network size on the slopes of KEGG and random networks. We observed negative correlation of network size with the slopes of KEGG and random networks (Pearson Correlation Coefficient (PCC) -0.58/-0.86, respectively, Figure 2, Supplementary Table S2). These observations confirm the notion that coherence estimation using internal vs. external degree distributions should be controlled for network size.

@@

Given the dependence of internal and external degree distributions on network size, we designed a simple strategy to account for it. Briefly, for each phenotype-specific network of a given size, we create size-matched references of high and low coherence. Specifically, we select the 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections) and 100 random networks matched in size for the corresponding network of phenotype-specific genes. Consequently, we use the median slopes of internal vs. external degree distributions created by these size-matched references to normalize the slopes of the phenotype-specific networks to the [0, 1] range (see Methods), where 0/1 correspond to random/high coherence, respectively. We found that the use of size-matched references indeed alleviates the dependency between coherence and network size (PCC = 0.29, Supplementary Figure S1) and allows for direct comparison of normalized coherences across phenotypes.

@@

Besides size dependency, coherence estimates depend on the choice of PPI database used to build networks and estimate the corresponding internal vs. external degree distributions. To evaluate consistency of coherence estimates, we evaluated pairwise correlation of normalized coherences obtained using different PPI databases. Correlation between coherence estimates obtained using "STRING" and "STRING filtered" was high (PCC = 0.77). Expectedly, coherence estimates using Biogrid database were less similar to that of "STRING" and "STRING filtered"; however, the overall correlation remained high (PCC = 0.56, Supplementary Figure S1). Notably, coherence estimates using "STRING filtered" database were most similar to that of "Biogrid" (PCC = 0.57), indicating that filtering step indeed removes "noisy" PPIs and improves quality of coherence estimates. Overall, these results demonstrate consistency in coherence estimates using different PPI databases.

@@

To alleviate the dependency of degree distributions from the size of phenotype-associated gene sets and the content of a selected PPI database, we normalized slopes of phenotype-specific networks to the range of slopes formed by degree distributions of highly coherent and random networks. Specifically, we select the 30 MSigDb networks (10 from KEGG, GOCC, and REACTOME collections, $\beta_{min}$ median slope) and 100 random networks ($\beta_{max}$ median slope) matched in size for the corresponding network of phenotype-specific genes. For each phenotype $i$ with slope $\beta_i$, we calculate the normalized slope $\beta_{i\ norm} = \frac{\beta_{i} - \beta_{max}}{\beta_{min}- \beta_{max}}$, which gives a measure of coherence on the range of [0,1]. Higher values indicate a higher level of coherence similar to that of MSigDb networks. These values represent normalized coherence and can be compared across phenotypes and PPI databases. Normalized coherence estimates for excellively large networks ("Obesity-related traits", 779 genes, "Height", 418 genes) were excluded from calculating correlation between coherences obtained using different PPI databases (Supplementary Figure S1).

@@

**Supplementary Figure S1. Normalization of coherence estimates alleviates its network size dependence.** Red/blue gradient and numbers represent Pearson correlation coefficients for each pairwise comparison of the number of SNPs, genes, and the normalized coherence estimates using the corresponding PPI databases.
